Opportunities for improvement in the care of adult trauma patients arriving

### 2025-02-16 manuscript.Rmd
la till fler rader

##Key results
We found that the most common type of OFI is clinical judgment error across almost all shock severities. Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows 63% reduced odds of OFI when compared against those with no shock.
The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.

##Interpretation
Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypotheses may explain this finding.
First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage. This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.
Another explanation is the potential influence of “reversed” survival bias. Patients with BE < -10 have high mortality rates (25), and for these patients who die, the care process is interrupted prematurely. This may induce generally fewer procedures or shorter care time for patients with deadly outcomes, which means less chance for errors, and therefore OFIs. This would result in reduced amount of OFIs in groups with higher mortality, which in this case would be class 4.  
Interpretation of adjusted model after inclusion of ISS
When looking at the distribution of the significant and insignificant results of the BE group (table 6), it can be observed that the significance reverses for the classes when adding ISS. One explanation is that the unadjusted positive correlation may be driven by a shared variability with ISS, as we can see significant positive association between ISS and OFI in our results. This positive association between OFI and ISS have also been shown in a study by Husssein et al (26). It is important to note that Hussein’s study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion criteria.
Another hypothetical explanation is multicollinearity, as both BE and ISS assess injury severity. Theoretically anatomical injury, bleeding and physiological response should be closely interconnected. A correlation between ISS and BD (inversed BE) or SBP was also shown in the two studies by Mutscher et al. (11,25). Therefore, adjusting for each other may explain the inversion of OR. However, it may also result in results that’s difficult to interpret or lack any meaning in a clinical context. This may also explain why class 2 and 3 are not significant but class 4 are in the adjusted analysis, as seen in study by Hussein et al (26), ISS does not correlate linearly to OFI, but we included ISS linearly in the model.    
However, when adding ISS to the SBP group, the significance of class 4 disappears, but the significance of classes 2 and 3 does not change, like in the case for BE group. The same can be said about the OR, it does change but not nearly as major as in the BE class. Especially for class 2, as the OR merely changes from 1.06 to 1.11 and then to 0.96. Class 3 OR changes more, from 0.92 to 0.90 to 0.47; but the OR difference is still < 0.5, whereof for the BE group and SBP class 4 the difference is always > 0.5. The reason may be due to the cutoffs of the classification system, as the SBP classification does not follow the same range of shock as the BE classification. The results of the study that BE classification was originating from (25), shows that BE class 1 had a mean SBP of 132.6 mmHg and class IV had 94.8 mmHg. As an SBP of 132.6 mmHg is at the higher range of normotensive, we were therefore not able to use it as an upper limit, but the 94.8 mmHg is close to our approximation of 90 mmHg for class 4. The SBP classification is therefore only somewhat connected to the BE classification at the lower limit e.g. class 4. Also worth noting is that the divider between our SBP class 2 and 3, are arbitrary and made to connect with the ATLS system to make up a four-class system, as the 90 mmHg and 110 mmHg are from Eastridge et al (23) and Oyetunji et al (24) which only defined 110 mmHg and 90 mmHg and nothing in between. This means that this divider may have made the classes too small and no longer scientific.

###Comparison to literature
Previous studies (7,27) which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgment error (7), or errors in management (27). In comparison, our results showed the most common OFI is clinical judgment error (36%), while inadequate resources were second (22%), and delay in treatment comes third (17%), for the whole study population. These differences may be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part of our results, while clinical judgment error takes up a larger part. No studies were found to address this hypothesis of survivorship bias for OFI types.
For patients defined as in shock (Class 2-4), the most common OFI was still clinical judgment error (BE 34.7%; SBP 35.3%). This result is similar when compared to D O’Reilly’s study (28), in which the most common type of opportunities for performance improvements (OPI) was one involved with decision-making at 63 out of 150 OPIs.
We did not find any publication addressing shock severity and OFI. We have therefore looked at studies addressing preventable negative care outcomes and assessed severity by other factors but only found one that investigated predicting factors and preventable death. This study (29) showed that there was a OR correlation of 14.1 between preventable death and SBP ≥ 80mmHg, when comparing to SBP < 80 mmHg. This differs from our SBP findings, as our adjusted results are not statistically significant, and the significant unadjusted results don’t have that extreme of OR and are positively correlated with severe and not mild shock. These differences are not strange, but rather understandable, as the studies are conducted at different hospitals, populations, outcomes (surviving vs death) and different classifications (the lowest of our 4 SBP classes is 90mmHg, while they only got 2 classes that separate at 80mmHg). However, if reasoned from a shock standpoint where both SBP and BE are part of, and compare this result with our BE results, they show a similar trend. Our BE results showed reduced odds of OFI when comparing the most severe and the no shock class. This implies that if changing the reference point to the most severe shock class, the result would be increased odds of OFI when compared to the no shock class. Therefore, our results show similar trend to their study (29) result, that showed lesser shock (higher SBP) had increased odds ratio of preventable death. 
One surprising result is that there is no significance for the SBP group in our adjusted analysis. When comparing to the results from Hussein et al. (26), the SBP results conform, as it also showed significance for the unadjusted results below 90 mmHg and no significance when adjusted with other factors including ISS. 
Due to a lack of research regarding shock/injury severity and OFI/preventable death, it is difficult to determine whether the results are reasonable or not. It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.

##Strengths and limitations
Strength of this study includes the relatively large cohort of 4919 patients spread over 9 continuous years. From a methodical standpoint, even though an established shock/hemorrhage classification system was not used, two clinical parameters were used in parallel. This includes two perspectives, one physiological and one chemical, to best describe shock and at the same time still using the original parameters the clinicians meet on a day-to-day basis.  
One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters. Due to a lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry, we were not able to use categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework (30–32). The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the proprietary classification we simplified and built our classification from.
The choice to not include the ATLS hemorrhage classification was due to four reasons. The first one is ATLS tenth edition lacks numerical values for vital parameters. The second one is that base deficit alone is superior to vital parameters as a predictive parameter for mortality (33). The third one is due to the nature of having multiple parameters, which makes it difficult to categorize a patient with vital parameters in two or more classes at the same time. The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.
Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE was not routinely measured and result in inconsistency for certain patient groups, which risks selection bias.
Also worth noting is that the sample size for most of the shock classes (class 2 to 4) are roughly around 100 to 200 patients, whereof those with positive OFI are as low as 5 patients. This may result in over/underestimation of the OR, as there is higher chance of randomness in smaller sample sizes, which results in broader confidence intervals which may result in more “extreme” average OR.

@@

###Future studies
The next step for future studies is to validate the findings and improve the methods from this study. As previously discussed, the association between anatomical classification ISS and physiological parameters BE and SBP, and how they correlate to OFI are complex and not necessarily clinically relevant. One way to avoid this problem is using a validated classification for hemorrhagic shock with both anatomical and physiological components, like ABC, TASH, and so on. 
One other problem when interpreting the results is the “reversed survivorship bias”. Due to the design of trauma care quality database, when and where (level of care) the OFI occurred during the length of stay is not recorded. A prospective study can clarify whether shock severity directly affects OFI or if length of stay is a confounding factor. By observing when and where OFIs most commonly occur, it can determine whether the hypothesis of length of stay is correct. This also aids in understanding the reasons behind the nonlinear relationship between ISS, BE or SBP and OFI, as there are currently only theories but no studies.

>>>>

##Key results
We found that the most common type of OFI is clinical judgment error across almost all shock severities. Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows 63% reduced odds of OFI when compared against those with no shock.

The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.

##Interpretation
Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypotheses may explain this finding.

First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage. This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.

Another explanation is the potential influence of “reversed” survival bias. Patients with BE < -10 have high mortality rates (25), and for these patients who die, the care process is interrupted prematurely. This may induce generally fewer procedures or shorter care time for patients with deadly outcomes, which means less chance for errors, and therefore OFIs. This would result in reduced amount of OFIs in groups with higher mortality, which in this case would be class 4.

###Interpretation of adjusted model after inclusion of ISS
When looking at the distribution of the significant and insignificant results of the BE group (table 6), it can be observed that the significance reverses for the classes when adding ISS. One explanation is that the unadjusted positive correlation may be driven by a shared variability with ISS, as we can see significant positive association between ISS and OFI in our results. This positive association between OFI and ISS have also been shown in a study by Husssein et al (26). It is important to note that Hussein’s study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion criteria.

Another hypothetical explanation is multicollinearity, as both BE and ISS assess injury severity. Theoretically anatomical injury, bleeding and physiological response should be closely interconnected. A correlation between ISS and BD (inversed BE) or SBP was also shown in the two studies by Mutscher et al. (11,25). Therefore, adjusting for each other may explain the inversion of OR. However, it may also result in results that’s difficult to interpret or lack any meaning in a clinical context. This may also explain why class 2 and 3 are not significant but class 4 are in the adjusted analysis, as seen in study by Hussein et al (26), ISS does not correlate linearly to OFI, but we included ISS linearly in the model.

However, when adding ISS to the SBP group, the significance of class 4 disappears, but the significance of classes 2 and 3 does not change, like in the case for BE group. The same can be said about the OR, it does change but not nearly as major as in the BE class. Especially for class 2, as the OR merely changes from 1.06 to 1.11 and then to 0.96. Class 3 OR changes more, from 0.92 to 0.90 to 0.47; but the OR difference is still < 0.5, whereof for the BE group and SBP class 4 the difference is always > 0.5. The reason may be due to the cutoffs of the classification system, as the SBP classification does not follow the same range of shock as the BE classification. The results of the study that BE classification was originating from (25), shows that BE class 1 had a mean SBP of 132.6 mmHg and class IV had 94.8 mmHg. As an SBP of 132.6 mmHg is at the higher range of normotensive, we were therefore not able to use it as an upper limit, but the 94.8 mmHg is close to our approximation of 90 mmHg for class 4. The SBP classification is therefore only somewhat connected to the BE classification at the lower limit e.g. class 4. Also worth noting is that the divider between our SBP class 2 and 3, are arbitrary and made to connect with the ATLS system to make up a four-class system, as the 90 mmHg and 110 mmHg are from Eastridge et al (23) and Oyetunji et al (24) which only defined 110 mmHg and 90 mmHg and nothing in between. This means that this divider may have made the classes too small and no longer scientific.

###Comparison to literature
Previous studies (7,27) which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgment error (7), or errors in management (27). In comparison, our results showed the most common OFI is clinical judgment error (36%), while inadequate resources were second (22%), and delay in treatment comes third (17%), for the whole study population. These differences may be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part of our results, while clinical judgment error takes up a larger part. No studies were found to address this hypothesis of survivorship bias for OFI types.

For patients defined as in shock (Class 2-4), the most common OFI was still clinical judgment error (BE 34.7%; SBP 35.3%). This result is similar when compared to D O’Reilly’s study (28), in which the most common type of opportunities for performance improvements (OPI) was one involved with decision-making at 63 out of 150 OPIs.
We did not find any publication addressing shock severity and OFI. We have therefore looked at studies addressing preventable negative care outcomes and assessed severity by other factors but only found one that investigated predicting factors and preventable death. This study (29) showed that there was a OR correlation of 14.1 between preventable death and SBP ≥ 80mmHg, when comparing to SBP < 80 mmHg. This differs from our SBP findings, as our adjusted results are not statistically significant, and the significant unadjusted results don’t have that extreme of OR and are positively correlated with severe and not mild shock. These differences are not strange, but rather understandable, as the studies are conducted at different hospitals, populations, outcomes (surviving vs death) and different classifications (the lowest of our 4 SBP classes is 90mmHg, while they only got 2 classes that separate at 80mmHg). However, if reasoned from a shock standpoint where both SBP and BE are part of, and compare this result with our BE results, they show a similar trend. Our BE results showed reduced odds of OFI when comparing the most severe and the no shock class. This implies that if changing the reference point to the most severe shock class, the result would be increased odds of OFI when compared to the no shock class. Therefore, our results show similar trend to their study (29) result, that showed lesser shock (higher SBP) had increased odds ratio of preventable death.

One surprising result is that there is no significance for the SBP group in our adjusted analysis. When comparing to the results from Hussein et al. (26), the SBP results conform, as it also showed significance for the unadjusted results below 90 mmHg and no significance when adjusted with other factors including ISS. 
Due to a lack of research regarding shock/injury severity and OFI/preventable death, it is difficult to determine whether the results are reasonable or not. It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.

##Strengths and limitations
Strength of this study includes the relatively large cohort of 4919 patients spread over 9 continuous years. From a methodical standpoint, even though an established shock/hemorrhage classification system was not used, two clinical parameters were used in parallel. This includes two perspectives, one physiological and one chemical, to best describe shock and at the same time still using the original parameters the clinicians meet on a day-to-day basis.

One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters. Due to a lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry, we were not able to use categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework (30–32). The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the proprietary classification we simplified and built our classification from.

The choice to not include the ATLS hemorrhage classification was due to four reasons. The first one is ATLS tenth edition lacks numerical values for vital parameters. The second one is that base deficit alone is superior to vital parameters as a predictive parameter for mortality (33). The third one is due to the nature of having multiple parameters, which makes it difficult to categorize a patient with vital parameters in two or more classes at the same time. The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.

Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE was not routinely measured and result in inconsistency for certain patient groups, which risks selection bias.
Also worth noting is that the sample size for most of the shock classes (class 2 to 4) are roughly around 100 to 200 patients, whereof those with positive OFI are as low as 5 patients. This may result in over/underestimation of the OR, as there is higher chance of randomness in smaller sample sizes, which results in broader confidence intervals which may result in more “extreme” average OR.

@@

###Future studies
The next step for future studies is to validate the findings and improve the methods from this study. As previously discussed, the association between anatomical classification ISS and physiological parameters BE and SBP, and how they correlate to OFI are complex and not necessarily clinically relevant. One way to avoid this problem is using a validated classification for hemorrhagic shock with both anatomical and physiological components, like ABC, TASH, and so on.

One other problem when interpreting the results is the “reversed survivorship bias”. Due to the design of trauma care quality database, when and where (level of care) the OFI occurred during the length of stay is not recorded. A prospective study can clarify whether shock severity directly affects OFI or if length of stay is a confounding factor. By observing when and where OFIs most commonly occur, it can determine whether the hypothesis of length of stay is correct. This also aids in understanding the reasons behind the nonlinear relationship between ISS, BE or SBP and OFI, as there are currently only theories but no studies.
